Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to US Midwest agroecosystems

### 2022-04-01 Manuscript/Manuscript.Rmd
jhjh

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome agronomic weed species in the United States. Palmer amaranth is prevalent in the Southern Great Plains and Southeastern United States, and its range is expanding northward through natural dispersal and human intervention. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and gender from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations in the US Midwest. Results demonstrated that the first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from the second cohort (established in July). The first Palmer amaranth cohort produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas the second Palmer amaranth cohort produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} in bareground, soybean and corn, respectively. Palmer amaranth height was most impacted when growing in corn, and averaged 85.2 cm tall in the first cohort, and 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Palmer amaranth gender was slightly influenced by day of year, weight and height. We documented a high degree of plasticity in Palmer amaranth, which will presumably favor its adaptation and expansion in cropping systems north of its current range. Therefore, preventing Palmer amaranth dispersal into new habitats is the most effective management strategy.

@@

Palmer amaranth's plasticity allows it to respond successfully to environmental changes. Palmer amaranth demonstrates a high degree of plasticity to light, temperature, water availability, and human management [@jha2010]. Palmer amaranth has an extended germination period throughout the growing season [@ward2013]. Germination of Palmer amaranth was triggered by 18 C soil temperature at 5 cm depth [@keeley1987], and optimal germination and biomass production occurred at 35/30 C day and night temperatures [@guo2003]. Palmer amaranth emergence can be influenced by tillage and the use of preemergence herbicides [@chahal2021], and may result in weed germination shifts within a population, as documented in *Bassia scoparia* [@sbatella2010]. In an experiment where Palmer amaranth was subjected to continuous water stress, it survived and produced at least 14000 seeds plant\textsuperscript{-1} [@chahal2018]. Seeds from Palmer amaranth growing with limited water conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. Growing conditions and management practices also influence Palmer amaranth sex dimorphism and flowering pattern [@korres2017; @rumpa2019].

Palmer amaranth's current global range includes agronomic cropland in Italy [@milani2021], soybean producing regions of Brazil and Argentina [@larran2017; @kupper2017], and the Cotton Belt [@garetson2019; @bagavathiannan2016] and southern Great Plains of the United States [@kumar2020, @crespo2016a]. However, its range appears to be expanding steadily, and in some cases rapidly, in the U.S. For example, in states with large Palmer amaranth infestations its range is gradually pushing north as it displaces other weeds in agronomic row crops (R Werle, personal observation), aided by wildlife [@farmer2017] and agricultural machinery [@ward2013]. But there are other examples where it has suddenly become a problem weed hundreds of miles away from known infestations. New infestations in Michigan, Indiana and North Dakota appear to have resulted from Palmer amaranth contaminated livestock feed. New infestations in Iowa and Minnesota were associated with planting contaminated pollinator habitat seed [@yu2021]. Palmer amaranth infestations have not been detected in Canada, but Palmer amaranth seeds were found in sweet potato slips imported into the country [@page2021]. Global warming will create new opportunities for Palmer amaranth invasion. Although agronomic crops in warm environments like Australia and Sub-Saharan Africa are currently at greatest risk for Palmer amaranth invasion, warming temperatures will reduce barriers that may have limited Palmer amaranths spread into cooler climates like Canada and Northern Europe [@kistner2018; @briscoerunquist2019].

There are many areas in the U.S. Corn Belt (North Central states) where Palmer amaranth is not yet established, and its potential adaptability is untested. Nonetheless, the rapid expansion of Palmer amaranth across the Northern U.S. is concerning and warrants investigations on its adaptability. We do know that  Palmer amaranth caused yield loss in Illinois soybean fields [@davis2015], but an Iowa study showed that Palmer amaranth was not as well adapted as waterhemp (*Amaranthus tuberculatus*) to conditions in that geography [@baker2021]. We also know that Palmer amaranth plants that establish shortly after row-crop planting have a much greater impact on crop yield than plants that emerge after the crop has produced several leaves [@macrae2013].  Understanding Palmer amaranth morphology and development under different agroecosystems and across a wide geography can enhance our knowledge of its adaptability, and may also aid in designing effective tactics to limit is range expansion and minimize its negative effects on row crops. The objective of this study was to investigate the flowering pattern, gender, biomass production, and height of Palmer amaranth cohorts growing in corn, soybean and bareground environments across five locations in the North Central United States.

@@

After transplanting, Palmer amaranth flowering was monitored until the end of the study. When a plant flowered, the day was recorded, plant gender was identified (male or female), plant height was measured from soil surface to the top of plant. Also, aboveground plant, and the aboveground plant organs were harvested, then oven dried at 65 C until a constant weight was reached, and dry biomass (g plant\textsuperscript{-1}) was recorded. Plants were harvested at flowering because Palmer amaranth infestations do not occur at the Wisconsin or Illinois research locations. In our study, all locations followed the methodology of plant harvest at flowering initiation, except in Grant, NE. At the Grant, NE, location, all plants from the first cohort were harvested on July 6, 2018 or July 6, 2019, and all plants from the second cohort were harvested on August 17, 2018 or July 31, 2019, regardless of flowering status.

@@

The model goodness of fit was 0.23, 0.32, 0.40 using pseudo R-squared test from McFadden, Cox and Snell, and Cragg and Uhler, respectively. The likelihood ratio test showed a p-value of < 0.00. The average marginal effects showed that Palmer amaranth growing in corn resulted in 14.8% fewer females plants (Table 2). Moreover, increasing a cumulative flowering unit doy increases the probability of having a female plant by 0.4% (Table 2 and Figure5A). A similar trend was observed for weight (Figure 6B) and height (Figure 6C), where the probability of being female increased 0.1% and 0.2% for each unit increase of weight and height, respectively (Table 2).

@@

Our study demonstrate that Palmer amaranth is well adapted to growing conditions throughout the Midwestern United States, and is simply limited in its range by seed dispersal (Davis et al. 2015). We also confirmed Palmer amaranth's extraordinary plasticity to adapt to different agroecosystems. For example, Palmer amaranth mimicked crop architecture in competing for light (Figure 7), where plants growing in corn allocated resources to height, while plants growing in bareground allocated resources to numerous branches. Plants growing in the absence of a crop produced the greatest biomass, responding to a greater abundance of light, nutrient and water resources (Figure 4A, 3B). These results support the argument that Palmer amaranth can quickly evolve life-history traits to adapt to different cultural practices, similar to observations in a Palmer amaranth response to nitrogen study [@bravo2018]. Our results highlight Palmer amaranth as a threat to field crops and breeding more competitive crop varieties is likely to select more competitive weed biotypes [@bravo2017].

```{r fig7, echo = FALSE, message=FALSE, fig.show = "hide"}
knitr::include_graphics("../data analysis/figures/Figure 7.png")
```

@@

Our study demonstrated that Palmer amaranth is adapted to grow on arable land throughout the Midwestern U.S. Palmer amaranth's range will continue to expand if current cropping practices are continued. As waterhemp and Palmer amaranth begin to share the same habitat, it will increase weed management complexity. Preventing Palmer amaranth seed dispersal must be a priority. Regional collaboration is necessary to slow the spread of this aggressive and adaptable weed. Where Palmer amaranth occurs, management tactics should focus on limiting Palmer amaranth establishment until row crops can shade late-emerging plants. Increasing the diversity of crops in rotation, varying row crop planting date, narrowing row width, and increasing residue cover through the use of cover crops are all tactics that can minimize the growth and seed production of Palmer amaranth, and improve the sustainability of North Central United States cropping systems.

@@

```{r, Figure-1, ref.label = "fig1", echo = FALSE, message = FALSE, fig.height=3, fig.width=6, fig.align='center', fig.cap='Field research locations in the United States north central region.', out.width = "160mm", out.height = "90mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

@@

```{r, Figure-7, ref.label = "fig7", echo = FALSE, message = FALSE, fig.height=9, fig.width=4, fig.align='center', fig.cap='MCO (first manuscript author, 180 cm) holds harvested Palmer amaranth plants at 40 days after first cohort transplanting (A) and 33 days after second cohort transplanting (B) time. From left to right in each image, Palmer amaranth growing in bareground, soybean and corn in Arlington, Wisconsin.', out.width = "150mm", out.height = "90mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome agronomic weed species in the United States. Palmer amaranth is prevalent in the Southern Great Plains and Southeastern United States, and its range is expanding northward through natural dispersal and human intervention. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted in 2018 and 2019 to investigate morphology, flowering and gender from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations in the Midwest United States. Results demonstrated that the first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from the second cohort (established in July). The first Palmer amaranth cohort produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas the second Palmer amaranth cohort produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} in bareground, soybean and corn, respectively. Palmer amaranth height was most impacted when growing in corn, and averaged 85.2 cm tall in the first cohort, and 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Palmer amaranth gender was slightly influenced by day of year, weight and height. We documented a high degree of plasticity in Palmer amaranth, which will presumably favor its adaptation and expansion in cropping systems north of its current range. Therefore, preventing Palmer amaranth dispersal into new habitats is the most effective management strategy.

@@

Palmer amaranth's plasticity allows it to respond successfully to environmental changes. Palmer amaranth demonstrates a high degree of plasticity to light, temperature, water availability, and human management [@jha2010]. Palmer amaranth has an extended germination period throughout the growing season [@ward2013]. Germination of Palmer amaranth was triggered by 18 \textdegree C soil temperature at 5 cm depth [@keeley1987], and optimal germination and biomass production occurred at 35/30 \textdegree C day and night temperatures [@guo2003]. Palmer amaranth emergence can be influenced by tillage and the use of preemergence herbicides [@chahal2021], and may result in weed germination shifts within a population, as documented in *Bassia scoparia* [@sbatella2010]. In an experiment where Palmer amaranth was subjected to continuous water stress, it survived and produced at least 14000 seeds plant\textsuperscript{-1} [@chahal2018]. Seeds from Palmer amaranth growing with limited water conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. Growing conditions and management practices also influence Palmer amaranth sex dimorphism and flowering pattern [@korres2017; @rumpa2019].

Palmer amaranth's current global range includes agronomic cropland in Italy [@milani2021], soybean producing regions of Brazil and Argentina [@larran2017; @kupper2017], and the Cotton Belt [@garetson2019; @bagavathiannan2016] and southern Great Plains of the United States [@kumar2020; @crespo2016a]. However, its range appears to be expanding steadily, and in some cases rapidly, in the United States. We hypothesize that in states with large Palmer amaranth infestations its range is gradually pushing north as it displaces other weeds in agronomic row crops, aided by wildlife [@farmer2017] and agricultural machinery [@ward2013]. But there are other examples where it has suddenly become a problem weed hundreds of miles away from known infestations. New infestations in Michigan, Indiana and North Dakota appear to have resulted from Palmer amaranth contaminated livestock feed. New infestations in Iowa and Minnesota were associated with planting contaminated pollinator habitat seed [@yu2021]. Palmer amaranth infestations have not been detected in Canada, but Palmer amaranth seeds were found in sweet potato slips imported into the country [@page2021]. Global warming will create new opportunities for Palmer amaranth invasion. Although agronomic crops in warm environments like Australia and Sub-Saharan Africa are currently at greatest risk for Palmer amaranth invasion, warming temperatures will reduce barriers that may have limited Palmer amaranths spread into cooler climates like Canada and Northern Europe [@kistner2018; @briscoerunquist2019].

There are many areas in the US Corn Belt (North Central states) where Palmer amaranth is not yet established, and its potential adaptability is untested. Nonetheless, the rapid expansion of Palmer amaranth across the Northern United States is concerning and warrants investigations on its adaptability. It has been shown that Palmer amaranth caused yield loss in Illinois soybean fields [@davis2015], but an Iowa study showed that Palmer amaranth was not as well adapted as waterhemp (*Amaranthus tuberculatus*) to conditions in that geography [@baker2021]. It also known that Palmer amaranth plants that establish shortly after row-crop planting have a much greater impact on crop yield than plants that emerge after the crop has produced several leaves [@macrae2013].  Understanding Palmer amaranth morphology and development under different agroecosystems and across a wide geography can enhance our knowledge of its adaptability, and may also aid in designing effective tactics to limit is range expansion and minimize its negative effects on row crops. The objective of this study was to investigate the flowering pattern, gender, biomass production, and height of Palmer amaranth cohorts growing in corn, soybean and bareground environments across five locations in the North Central United States.

@@

After transplanting, Palmer amaranth flowering was monitored until the end of the study. When a plant flowered, the day was recorded, plant gender was identified (male or female), plant height was measured from soil surface to the top of plant. Also, aboveground plant, and the aboveground plant organs were harvested, then oven dried at 65 \textdegree C until a constant weight was reached, and dry biomass (g plant\textsuperscript{-1}) was recorded. Plants were harvested at flowering because Palmer amaranth is not established at the Wisconsin or Illinois research locations. In our study, all locations followed the methodology of plant harvest at flowering initiation, except in Grant, NE. At the Grant, NE, location, all plants from the first cohort were harvested on July 6, 2018 or July 6, 2019, and all plants from the second cohort were harvested on August 17, 2018 or July 31, 2019, regardless of flowering status.

@@

The model goodness of fit was 0.23, 0.32, 0.40 using pseudo R-squared test from McFadden, Cox and Snell, and Cragg and Uhler, respectively. The likelihood ratio test showed a p-value of 0.00. The average marginal effects showed that Palmer amaranth growing in corn resulted in 14.8% fewer females plants (Table 2). Moreover, increasing a cumulative flowering unit doy increases the probability of having a female plant by 0.4% (Table 2 and Figure5A). A similar trend was observed for weight (Figure 6B) and height (Figure 6C), where the probability of being female increased 0.1% and 0.2% for each unit increase of weight and height, respectively (Table 2).

@@

Our study demonstrate that Palmer amaranth is well adapted to growing conditions throughout the Midwestern US, and is simply limited in its range by seed dispersal (Davis et al. 2015). We also confirmed Palmer amaranth's extraordinary plasticity to adapt to different agroecosystems. For example, Palmer amaranth mimicked crop architecture in competing for light, where plants growing in corn allocated resources to height, while plants growing in bareground allocated resources to numerous branches. Plants growing in the absence of a crop produced the greatest biomass, responding to a greater abundance of light, nutrient and water resources (Figure 4A, 3B). These results support the argument that Palmer amaranth can quickly evolve life-history traits to adapt to different cultural practices, similar to observations in a Palmer amaranth response to nitrogen study [@bravo2018]. Our results highlight Palmer amaranth as a threat to field crops and breeding more competitive crop varieties is likely to select more competitive weed biotypes [@bravo2017].

```{r fig7, echo = FALSE, message=FALSE, fig.show = "hide", include = FALSE}
knitr::include_graphics("../data analysis/figures/Figure 7.png")
```

@@

Our study demonstrated that Palmer amaranth is adapted to grow on arable land throughout the north central United States. Palmer amaranth's range will continue to expand if current cropping practices are continued. As waterhemp and Palmer amaranth begin to share the same habitat, it will increase weed management complexity. Preventing Palmer amaranth seed dispersal must be a priority. Regional collaboration is necessary to slow the spread of this aggressive and adaptable weed. Where Palmer amaranth occurs, management tactics should focus on limiting Palmer amaranth establishment until row crops can shade late-emerging plants. Increasing the diversity of crops in rotation, varying row crop planting date, narrowing row width, and increasing residue cover through the use of cover crops are all tactics that can minimize the growth and seed production of Palmer amaranth, and improve the sustainability of US North Central cropping systems.

@@

```{r, Figure-1, ref.label = "fig1", echo = FALSE, message = FALSE, fig.height=3, fig.width=6, fig.align='center', fig.cap='Field research locations in the north central United States region.', out.width = "160mm", out.height = "90mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

@@

```{r, Figure-7, include=FALSE, ref.label = "fig7", echo = FALSE, message = FALSE, fig.height=9, fig.width=4, fig.align='center', fig.cap='MCO (first manuscript author, 180 cm) holds harvested Palmer amaranth plants at 40 days after first cohort transplanting (A) and 33 days after second cohort transplanting (B) time. From left to right in each image, Palmer amaranth growing in bareground, soybean and corn in Arlington, Wisconsin.', out.width = "150mm", out.height = "90mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

### 2022-04-01 Manuscript/Manuscript.tex
jhjh

Palmer amaranth's plasticity allows it to respond successfully to
environmental changes. Palmer amaranth demonstrates a high degree of
plasticity to light, temperature, water availability, and human
management (Jha et al., 2010). Palmer amaranth has an extended
germination period throughout the growing season (Ward et al., 2013).
Germination of Palmer amaranth was triggered by 18 C soil temperature at
5 cm depth (Keeley et al., 1987), and optimal germination and biomass
production occurred at 35/30 C day and night temperatures (Guo and
Al-Khatib, 2003). Palmer amaranth emergence can be influenced by tillage
and the use of preemergence herbicides (Chahal et al., 2021), and may
result in weed germination shifts within a population, as documented in
\emph{Bassia scoparia} (Sbatella and Wilson, 2010). In an experiment
where Palmer amaranth was subjected to continuous water stress, it
survived and produced at least 14000 seeds plant\textsuperscript{-1}
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019).

Palmer amaranth's current global range includes agronomic cropland in
Italy (Milani et al., 2021), soybean producing regions of Brazil and
Argentina (Larran et al., 2017; Küpper et al., 2017), and the Cotton
Belt (Garetson et al., 2019; Bagavathiannan and Norsworthy, 2016) and
southern Great Plains of the United States Crespo et al. (2016).
However, its range appears to be expanding steadily, and in some cases
rapidly, in the U.S. For example, in states with large Palmer amaranth
infestations its range is gradually pushing north as it displaces other
weeds in agronomic row crops (R Werle, personal observation), aided by
wildlife (Farmer et al., 2017) and agricultural machinery (Ward et al.,
2013). But there are other examples where it has suddenly become a
problem weed hundreds of miles away from known infestations. New
infestations in Michigan, Indiana and North Dakota appear to have
resulted from Palmer amaranth contaminated livestock feed. New
infestations in Iowa and Minnesota were associated with planting
contaminated pollinator habitat seed (Yu et al., 2021). Palmer amaranth
infestations have not been detected in Canada, but Palmer amaranth seeds
were found in sweet potato slips imported into the country (Page et al.,
2021). Global warming will create new opportunities for Palmer amaranth
invasion. Although agronomic crops in warm environments like Australia
and Sub-Saharan Africa are currently at greatest risk for Palmer
amaranth invasion, warming temperatures will reduce barriers that may
have limited Palmer amaranths spread into cooler climates like Canada
and Northern Europe (Kistner and Hatfield, 2018; Briscoe Runquist et
al., 2019).

There are many areas in the U.S. Corn Belt (North Central states) where
Palmer amaranth is not yet established, and its potential adaptability
is untested. Nonetheless, the rapid expansion of Palmer amaranth across
the Northern U.S. is concerning and warrants investigations on its
adaptability. We do know that Palmer amaranth caused yield loss in
Illinois soybean fields (Davis et al., 2015), but an Iowa study showed
that Palmer amaranth was not as well adapted as waterhemp
(\emph{Amaranthus tuberculatus}) to conditions in that geography (Baker,
2021). We also know that Palmer amaranth plants that establish shortly
after row-crop planting have a much greater impact on crop yield than
plants that emerge after the crop has produced several leaves (MacRae et
al., 2013). Understanding Palmer amaranth morphology and development
under different agroecosystems and across a wide geography can enhance
our knowledge of its adaptability, and may also aid in designing
effective tactics to limit is range expansion and minimize its negative
effects on row crops. The objective of this study was to investigate the
flowering pattern, gender, biomass production, and height of Palmer
amaranth cohorts growing in corn, soybean and bareground environments
across five locations in the North Central United States.

@@

\hypertarget{field-study}{%
\subsection*{Field study}\label{field-study}}
\addcontentsline{toc}{subsection}{Field study}

The experiment was conducted in 2018 and 2019 under field conditions at
five locations: Arlington, WI, Clay Center, NE,, Grant, NE, Lincoln, NE,
and Macomb, IL.

Fields were conventionally tilled prior to crop planting. Corn and
soybean were planted in 76-cm row spacing (Table 1). Monthly mean air
temperature and total precipitation were obtained using Daymet weather
data from June through September across the five locations in 2018 and
2019 (Correndo et al., 2021) (Figure 2).

@@

\caption{\label{tab:unnamed-chunk-2}Field study attributes from Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL.}
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lllllll}
\toprule
Attributes &  & Arlington, WI & Clay Center, NE & Grant, NE & Lincoln, NE & Macomb, IL\\
\midrule
Bareground &  &  &  &  &  & \\

 & Weed control\textsuperscript{a} & \emph{S}-metolachor\textsuperscript{b} fb glyphosate\textsuperscript{c}  & saflufenacil + imazethapyr + pyroxasulfone\textsuperscript{d} & glyphosate\textsuperscript{c} & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & glyphosate + \emph{S}-metolachor\textsuperscript{f} \\

@@

 & Weed control & \emph{S}-metolachor fb glyphosate & \emph{S}-metolachlor + trazine + mesotrione + bicyclopyrone\textsuperscript{e} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & glyphosate + \emph{S}-metolachor \\

@@

 & Type & Plano-silt-loam & Crete Silt Loam & 2018: Mace silt loam / 2019: Kuma silt loam & Crete silty clay loam & \\

 & pH & 6.6 & 6.5 & 2018: 6.1 / 2019: 5.1 & 5.2 & \\

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) & 3.5 & 3 & 2018: 2 / 2019: 1.7 & 3.4 & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a}Hoe weeding; \textsuperscript{b}\emph{S}--metolachor, 1324 g ai ha\textsuperscript{-1}; \textsuperscript{c}glyphoste, 1262 g ae ha\textsuperscript{-1}; \textsuperscript{d}saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha\textsuperscript{-1}; \textsuperscript{e}\emph{S}-metolachlor + trazine + mesotrione, + bicyclopyrone, 2409 g ai ha\textsuperscript{-1}; \textsuperscript{f}\emph{S}--metolachor, 1070 g ai ha\textsuperscript{-1}; \textsuperscript{g}glyphoste, 870 g ae ha\textsuperscript{-1}} \\
\end{tabular}}
\end{table}

@@

The statistical analyses were performed using R statistical software
version 4.0.1 (R Core Team, 2021).

@@

The model goodness of fit was 0.23, 0.32, 0.40 using pseudo R-squared
test from McFadden, Cox and Snell, and Cragg and Uhler, respectively.
The likelihood ratio test showed a p-value of \textless{} 0.00. The
average marginal effects showed that Palmer amaranth growing in corn
resulted in 14.8\% fewer females plants (Table 2). Moreover, increasing
a cumulative flowering unit doy increases the probability of having a
female plant by 0.4\% (Table 2 and Figure5A). A similar trend was
observed for weight (Figure 6B) and height (Figure 6C), where the
probability of being female increased 0.1\% and 0.2\% for each unit
increase of weight and height, respectively (Table 2).

@@

Our study demonstrate that Palmer amaranth is well adapted to growing
conditions throughout the Midwestern United States, and is simply
limited in its range by seed dispersal (Davis et al.~2015). We also
confirmed Palmer amaranth's extraordinary plasticity to adapt to
different agroecosystems. For example, Palmer amaranth mimicked crop
architecture in competing for light (Figure 7), where plants growing in
corn allocated resources to height, while plants growing in bareground
allocated resources to numerous branches. Plants growing in the absence
of a crop produced the greatest biomass, responding to a greater
abundance of light, nutrient and water resources (Figure 4A, 3B). These
results support the argument that Palmer amaranth can quickly evolve
life-history traits to adapt to different cultural practices, similar to
observations in a Palmer amaranth response to nitrogen study (Bravo et
al., 2018). Our results highlight Palmer amaranth as a threat to field
crops and breeding more competitive crop varieties is likely to select
more competitive weed biotypes (Bravo et al., 2017).

@@

\leavevmode\hypertarget{ref-rcoreteam2021}{}%
R Core Team (2021). \emph{R: {The R Project} for {Statistical
Computing}}. {Vienna, Austria}: {R Foundation for Statistical Computing}
Available at: \url{https://www.r-project.org/} {[}Accessed August 24,
2021{]}.

@@

\caption{Field research locations in the United States North Central region.}\label{fig:Figure-1}
\end{figure}

@@

{\centering \includegraphics[width=150mm,height=90mm]{../data analysis/figures/Figure 7}

\caption{MCO (first manuscript author, 180 cm) holds harvested Palmer amaranth plants at 40 days after first cohort transplanting (A) and 33 days after second cohort transplanting (B) time. From left to right in each image, Palmer amaranth growing in bareground, soybean and corn in Arlington, Wisconsin.}\label{fig:Figure-7}
\end{figure}

>>>>

Palmer amaranth's plasticity allows it to respond successfully to
environmental changes. Palmer amaranth demonstrates a high degree of
plasticity to light, temperature, water availability, and human
management (Jha et al., 2010). Palmer amaranth has an extended
germination period throughout the growing season (Ward et al., 2013).
Germination of Palmer amaranth was triggered by 18 \textdegree C soil
temperature at 5 cm depth (Keeley et al., 1987), and optimal germination
and biomass production occurred at 35/30 \textdegree C day and night
temperatures (Guo and Al-Khatib, 2003). Palmer amaranth emergence can be
influenced by tillage and the use of preemergence herbicides (Chahal et
al., 2021), and may result in weed germination shifts within a
population, as documented in \emph{Bassia scoparia} (Sbatella and
Wilson, 2010). In an experiment where Palmer amaranth was subjected to
continuous water stress, it survived and produced at least 14000 seeds
plant\textsuperscript{-1} (Chahal et al., 2018). Seeds from Palmer
amaranth growing with limited water conditions were heavier, less
dormant, and prompt for germination (Matzrafi et al., 2021). Growing
conditions and management practices also influence Palmer amaranth sex
dimorphism and flowering pattern (Korres et al., 2017; Rumpa et al.,
2019).

Palmer amaranth's current global range includes agronomic cropland in
Italy (Milani et al., 2021), soybean producing regions of Brazil and
Argentina (Larran et al., 2017; Küpper et al., 2017), and the Cotton
Belt (Garetson et al., 2019; Bagavathiannan and Norsworthy, 2016) and
southern Great Plains of the United States (Kumar et al., 2020; Crespo
et al., 2016). However, its range appears to be expanding steadily, and
in some cases rapidly, in the United States. We hypothesize that in
states with large Palmer amaranth infestations its range is gradually
pushing north as it displaces other weeds in agronomic row crops, aided
by wildlife (Farmer et al., 2017) and agricultural machinery (Ward et
al., 2013). But there are other examples where it has suddenly become a
problem weed hundreds of miles away from known infestations. New
infestations in Michigan, Indiana and North Dakota appear to have
resulted from Palmer amaranth contaminated livestock feed. New
infestations in Iowa and Minnesota were associated with planting
contaminated pollinator habitat seed (Yu et al., 2021). Palmer amaranth
infestations have not been detected in Canada, but Palmer amaranth seeds
were found in sweet potato slips imported into the country (Page et al.,
2021). Global warming will create new opportunities for Palmer amaranth
invasion. Although agronomic crops in warm environments like Australia
and Sub-Saharan Africa are currently at greatest risk for Palmer
amaranth invasion, warming temperatures will reduce barriers that may
have limited Palmer amaranths spread into cooler climates like Canada
and Northern Europe (Kistner and Hatfield, 2018; Briscoe Runquist et
al., 2019).

There are many areas in the US Corn Belt (North Central states) where
Palmer amaranth is not yet established, and its potential adaptability
is untested. Nonetheless, the rapid expansion of Palmer amaranth across
the Northern United States is concerning and warrants investigations on
its adaptability. It has been shown that Palmer amaranth caused yield
loss in Illinois soybean fields (Davis et al., 2015), but an Iowa study
showed that Palmer amaranth was not as well adapted as waterhemp
(\emph{Amaranthus tuberculatus}) to conditions in that geography (Baker,
2021). It also known that Palmer amaranth plants that establish shortly
after row-crop planting have a much greater impact on crop yield than
plants that emerge after the crop has produced several leaves (MacRae et
al., 2013). Understanding Palmer amaranth morphology and development
under different agroecosystems and across a wide geography can enhance
our knowledge of its adaptability, and may also aid in designing
effective tactics to limit is range expansion and minimize its negative
effects on row crops. The objective of this study was to investigate the
flowering pattern, gender, biomass production, and height of Palmer
amaranth cohorts growing in corn, soybean and bareground environments
across five locations in the North Central United States.

@@

@@

\caption{\label{tab:unnamed-chunk-2}Field study attributes from Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL}
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lllllll}
\toprule
Attributes &  & Arlington, WI & Clay Center, NE & Grant, NE & Lincoln, NE & Macomb, IL\\
\midrule
Bareground &  &  &  &  &  & \\

 & Weed control\textsuperscript{a} & \emph{S}-metolachor\textsuperscript{b} fb glyphosate\textsuperscript{c} /  & saflufenacil + imazethapyr + pyroxasulfone\textsuperscript{d} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & glyphosate + \emph{S}-metolachor\textsuperscript{f} \\

@@

 & Weed control & \emph{S}-metolachor fb glyphosate & \emph{S}-metolachlor + atrazine + mesotrione + bicyclopyrone\textsuperscript{e} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & glyphosate + \emph{S}-metolachor \\

@@

 & Type & Plano-silt-loam & Crete Silt Loam & 2018: Mace silt loam / 2019: Kuma silt loam & Crete silty clay loam &  2018: Osco silt loam / 2019: Keomah silt loam \\

 & pH & 6.6 & 6.5 & 2018: 6.1 / 2019: 5.1 & 5.2 &  2018: 6.8 / 2019: 7.5\\

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) & 3.5 & 3 & 2018: 2 / 2019: 1.7 & 3.4 &  2018: 2.0 / 2019: 1.3\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a}Hoe weeding; \textsuperscript{b}\emph{S}--metolachor, 1324 g ai ha\textsuperscript{-1}; \textsuperscript{c}glyphoste, 1262 g ae ha\textsuperscript{-1}; \textsuperscript{d}saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha\textsuperscript{-1}; \textsuperscript{e}\emph{S}-metolachlor + atrazine + mesotrione, + bicyclopyrone, 2409 g ai ha\textsuperscript{-1}; \textsuperscript{f}\emph{S}--metolachor, 1070 g ai ha\textsuperscript{-1}; \textsuperscript{g}glyphoste, 870 g ae ha\textsuperscript{-1}} \\
\end{tabular}}
\end{table}

@@

The statistical analyses were performed using R statistical software
version 4.0.1 (Team, 2021).

@@

The model goodness of fit was 0.23, 0.32, 0.40 using pseudo R-squared
test from McFadden, Cox and Snell, and Cragg and Uhler, respectively.
The likelihood ratio test showed a p-value of 0.00. The average marginal
effects showed that Palmer amaranth growing in corn resulted in 14.8\%
fewer females plants (Table 2). Moreover, increasing a cumulative
flowering unit doy increases the probability of having a female plant by
0.4\% (Table 2 and Figure5A). A similar trend was observed for weight
(Figure 6B) and height (Figure 6C), where the probability of being
female increased 0.1\% and 0.2\% for each unit increase of weight and
height, respectively (Table 2).

@@

Our study demonstrate that Palmer amaranth is well adapted to growing
conditions throughout the Midwestern US, and is simply limited in its
range by seed dispersal (Davis et al.~2015). We also confirmed Palmer
amaranth's extraordinary plasticity to adapt to different
agroecosystems. For example, Palmer amaranth mimicked crop architecture
in competing for light, where plants growing in corn allocated resources
to height, while plants growing in bareground allocated resources to
numerous branches. Plants growing in the absence of a crop produced the
greatest biomass, responding to a greater abundance of light, nutrient
and water resources (Figure 4A, 3B). These results support the argument
that Palmer amaranth can quickly evolve life-history traits to adapt to
different cultural practices, similar to observations in a Palmer
amaranth response to nitrogen study (Bravo et al., 2018). Our results
highlight Palmer amaranth as a threat to field crops and breeding more
competitive crop varieties is likely to select more competitive weed
biotypes (Bravo et al., 2017).

@@

\leavevmode\hypertarget{ref-rcoreteam2021}{}%
Team, R. C. (2021). \emph{R: {The R Project} for {Statistical
Computing}}. {Vienna, Austria}: {R Foundation for Statistical Computing}
Available at: \url{https://www.r-project.org/} {[}Accessed August 24,
2021{]}.

@@

\caption{Field research locations in the north central United States region.}\label{fig:Figure-1}
\end{figure}

@@
